Implications of afforestation policies for global markets

### 2024-01-22 paper.Rmd
Initial edits

---
title: "Implications of afforestation policies for global markets"
date: "`r format(Sys.time(), '%B %d, %Y')`"
author: "Eric Davis, Maros Ivanic, Brent Sohngen"
output: bookdown::pdf_document2
toc: FALSE
bibliography: references.bib
header-includes:
    - \usepackage{setspace}\doublespacing
abstract: "The projected growth of the population and incomes as implied by various shared socioeconomic pathways is expected to create  pressures on converting forests into farmland as food prices increase. Depending on the specific path taken, SSP1 through SSP5, and the continuation of the current trends in agricultural productivity growth, the market impacts are expected to vary but generally result in increases in food prices. Policies aimed at reversing the deforestation trend by afforesting cropland would result in severe negative market impacts such as  higher food prices, especially in low agricultural productivity/high population growth scenarios."
---

@@

In 2020, there were 4.1 billion hectares of forestland globally covering roughly 31.4% of the land [@fao2023]. These forests play an important role in maintaining and growing the land-based carbon sink by  sequestering about 30% of GHG emissions currently [@roe2019]. This is because of forests proven ability to sequester carbon not only in the stems of trees [@davis2022] but in their soils as well [@guo2021]. With increasingly negative climate impacts, the pressure to enhance the terrestrial sink is growing, and one of the tools to achieve this is afforestation. As such, there is growing attention being placed globally to afforestation efforts, which is highlighted by the UN's declaration of the years from 2021 through 2030 as the UN Decade of Ecosystem Restoration [@sewell2020]. Many of the Nationally Determined Contributions (NDCs) countries have submitted, which detail their plans to keep warming under the 2C threshold, also rely to a significant degree on afforestation projects [@roe2019].

The global gains that can be realized from afforestation, however, are uncertain, with broad estimates of cumulative sequestration ranging from 42 gigatonnes to 700 gigatonnes by the end of the century [@veldman2019]  [@bastin201]  [@Humpenöder2014]  [@sathaye2006]  [@doelman2020]. In addition, the amount of land conversion that would be required also varies greatly, with estimates ranging from 0.3 billion hectares to 2.8 billion hectares [@veldman2019]  [@bastin201]  [@Humpenöder2014]  [@doelman2020]. In many cases, these results suggest massive losses in the amount of pasture and cropland that would be available. This puts into question the feasibility of attaining these results, as almost 90 percent of deforestation worldwide is undertaken to expand the amount of agricultural land [@fao2021]. There are many reasons for this. For example, as nations develop and disposable incomes rise, consumers often change to more meat-based diets. Population growth is also a driver and when it is especially high, which is the case at present throughout most of Africa, this also puts pressure on policymakers to allow greater conversion of forestland into farmland. Still, previous research has suggested that substantive gains in the carbon sink (+23.8PgCo2e/yr) can be made without sacrificing agricultural land [@griscom2017].

Faced with competing land-use demands, afforestation efforts need to be targeted to the regions of the world where they will provide the most benefits. Afforestation in boreal zones has been shown to provide the smallest net benefits, as the carbon sequestrations benefits of conifer forests are mitigated by the reduction in albedo in winter [@mykleby2017] [@li2015]. Coniferous forests have also been shown to increase soil organic carbon less than broadleaf forests [@guo2021]. Conversely, tropical forests have been linked with a strong positive result from afforestation, as the carbon sequestration benefit is complemented by a cooling impact due to changes in both albedo and evapotranspiration, and with temperate forests, the impact of afforestation is also shown to be positive, albeit less strongly so due to a winter warming effect [@li2015].

South America and Sub-Saharan Africa are two regions that show much potential for afforestation. It is estimated that these two regions hold at least 50% of the potential global gain while regions such as Northern Africa and the Middle East show little promise as forest growth rates are quite low [@doelman2020]. Thus, it is promising that the African Forest Landscape Restoration Initiative (AFR100), which aims to afforest 100 million hectares by 2030, has roughly 70% of its commitments coming from Sub-Saharan nations [@auda2023]. While many of these nations may be well-suited in biophysical terms for afforestation, there is concern about how long-lasting any investment may be due to proximate zones of political instability. There are other nations that have a big potential for gains from afforestation, but they also face challenges. India, for example, has a great potential to increase its carbon sequestration through afforestation, but due to its massive population and food security worries, its NDC does not entertain the idea of agricultural land being repurposed and instead primarily considers targeting wastelands that may not be well suited for afforestation [@amjath2019]. Moreover, as land-use decisions in one country also tend to have impacts that spillover into many others [@meyfroidt2010] [@lambin2011], analysis needs to be conducted at a global scale.

Steinbuks and Hertel [@steinbuks2012] developed the FABLE model (Forestry, Agriculture, Biofuels, Land Use, and Environment) as a global partial equilibrium model to analyze optimal land use within a context of increasing demand for food, bioenergy, forest products, and non-provisioning ecosystem services, while also meeting greenhouse gas targets. The model encompasses nine sectors, including agriculture, livestock, food processing, biofuels, energy, forest, timber processing, and ecosystem services. While FABLE is a powerful tool for examining the optimal trajectory of various land uses under specific demand assumptions, it primarily focuses on specific sectors of the economy, neglecting general equilibrium effects and socio-economic or household welfare impacts [@steinbuks2012] [@hertel2017]. Other efforts have used the Global Trade Analysis Project (GTAP) computable general equilibrium model to address such questions, but its standard model does not well account for land-use changes. As a result, the Land Use and Land Cover Database within the GTAP framework, which incorporates forestry remote-sensing products, was created [@baldos2012]. It was further strengthened as part of the GTAP Agro-Ecological Zone (GTAP-AEZ) model, with the use of spatially explicit global land use data and through the incorporation of intra- and inter-regional land and land-based greenhouse emissions heterogeneity [@hertel2008]. While the GTAP-AEZ model benefits from the inclusion of competition among different crops, grazing, and forest-based uses, it has limitations, primarily due to its infrequent updates. The KLUM@GTAP framework integrated the Kleines Land Use Model (KLUM) with an extended yet static version of GTAP called GTAP-EFL to evaluate the impacts of climate change on cropland allocation. GTAP-EFL separates energy factors from intermediate inputs and incorporates them into capital, while also considering CO2 emissions. KLUM, a global agricultural land-use model, connects the economy to global crop allocation to maximize producer returns under specific risk assumptions. KLUM@GTAP substitutes the land allocation mechanism within GTAP-EFL by utilizing regionally aggregated area changes in cropland determined by KLUM to update cropland shares in GTAP-EFL [@ronneberger2009].

More recent modeling attempts of land-use change have begun to move away from comparative static analysis. For example, the dynamic GTAP model, supported by the GTAP-AEZ database, considers land market effects, which were identified as significant in driving results by Stevenson et al. [@stevenson2013], and has been utilized to evaluate the impacts of crop intensification on land use. It, however, has been critiqued, as global aggregates may mask localized shifts that can have implications for ecologically significant areas [@byerlee2014]. This project will attempt to get around these concerns and assess the impact of various likely afforestation scenarios by melding GTAP with the Global Timber Model (GTM), which is a dynamic optimization model that can examine the impact of afforestation or avoided deforestation on the carbon sink [@sohngen2001]. By joining these two models and passing outputs back and forth between the models, any limitations of either model can be minimized and more robust results, therefore, can be attained.

@@

1) Full attainment of AFR100 pledges as part of Bonn Challenge: This says that there will be a 127.81 million hectare increase in forest area just in Sub-Saharan Africa by 2030 [@bonn2023]. To see the list of countries and the hectares pledged, please view Scenario 1 of the Excel document "Bonn Challenge-Africa".
2) Partial attainment of AFR100 pledges as part of Bonn Challenge: This scenario analyzes full attainment for nations with conflict severity level less than or equal to limited and 0% attainment for regions with a higher conflict severity rating. Using this measure, there would only be a 54.95 million hectare increase in forest area by 2030 [@bonn2023]  [@acled2023]. To see the list of countries and the hectares pledged, please view Scenario 2 of the Excel document "Bonn Challenge-Africa".
3) Full attainment of Latin America pledges as part of Bonn Challenge: This says that there will be a 33.64 million hectare increase in forest area by 2030 [@bonn2023]. To see the list of countries and the hectares pledged, please view the Latin America tab of the Excel document "Bonn Challenge-Other Regions".
4) Full attainment of tropical nations' pledges as part of Bonn Challenge: This says that there will be a 182.30 million hectare increase in forest area by 2030 [@bonn2023]. To see the list of countries and the hectares pledged, please view the Tropical Forest tab of the Excel document "Bonn Challenge-Other Regions". Nations that lie in both tropical and sub-tropical zones and taken to be tropical.
5) Full attainment of UN strategic plan for forests 2030 (UNFF): This says that all deforestation will halt and there will be a 3% (~120 million hectare) increase in forest area worldwide by 2030 [@sewell2020]. This will be modeled as a uniform 3% increase in forested area across all nations.
6) Expansion of UN strategic plan for forests: This says that all deforestation will halt and there will be a 15% (~600 million hectare) increase in forest area worldwide by 2030 [@sewell2020]. This will be modeled as a uniform 15% increase in forested area across all nations.

DeFries et al. [@defries2010] found a strong correlation between forest loss and exports of agricultural products. On the other hand, Turner et al. [@turner2005] investigated the role of trade on forests in the context of the Free Trade Area of the Americas (FTAA) using the spatial partial-equilibrium Global Forest Products Model and found that FTAA has a positive impact on forest resources.
# References

>>>>

---
title: "Implications of afforestation policies for global markets"
date: "`r format(Sys.time(), '%B %d, %Y')`"
author: "Eric Davis, Maros Ivanic, Brent Sohngen"
output: 
  bookdown::pdf_document2:
    keep_tex: TRUE
toc: FALSE
bibliography: references.bib
header-includes:
    - \usepackage{setspace}\doublespacing
abstract: "The projected growth of the population and incomes as implied by various shared socioeconomic pathways is expected to create  pressures on converting forests into farmland as food prices increase. Depending on the specific path taken, SSP1 through SSP5, and the continuation of the current trends in agricultural productivity growth, the market impacts are expected to vary but generally resulting in increases in food prices. Policies aimed at reversing the deforestation trend by afforesting crop land would result in severe negative market impacts such as  higher food prices, especially in low agriculutrla productivity/high population growht scenarios."
editor_options: 
  chunk_output_type: console
---

@@

In 2020, there were 4.1 billion hectares of forestland globally that cover roughly 31.4% of the land [@fao2023]. These forests play an important role in maintaining and growing the land-based carbon sink, currently sequestering about 30% of GHG emmissions [@roe2019]. With increasingly negative climate impacts, the pressure to enhance the terrestrial sink is growing, and one of the tools to achieve this is afforestation. There is growing attention being placed globally to afforestation efforts, which is highligted by the UN's declaration of the years from 2021 through 2030 as the UN Decade of Ecosystem Resotration [@sewell2020], and many of the Nationally Determined Contributions (NDCs) countries have submitted, which detail their plans to keep warming under the 2C threshold, rely to a significant degree on afforestation projects [@roe2019]. This is because of forests proven ability to sequester carbon not only in the stems of trees [@davis2022] but in their soils as well [@guo2021].

The global gains that can be realized from afforestion, however, are uncertain, with broad estimates of cumulative sequestration ranging from 42 gigatonnes to 700 gigatonnes by the end of the century [@veldman2019]  [@bastin201]  [@humpenoder2014]  [@sathaye2006]  [@doelman2020]. In addition, the amount of land conversion that would be required also varies greatly, with estimates ranging from 0.3 billion hectares to 2.8 billion hectares [@veldman2019]  [@bastin201]  [@humpenoder2014]  [@doelman2020].  In many cases, these results suggest massive losses in pasture and cropland. This puts into question the feasibility of attaining these results, as forests are continually threatened by issues such as rising incomes and food prices and population growth that put pressure on policymakers to allow greater conversion of forestland into farmland. Previous research though has suggested that substantive gains in the carbon sink (+23.8PgCo2e/yr) can be made without sacrificing agricultural land [@griscom2017]. Still, the manner in which these various and oftentimes competing concerns are addressed will have significant impacts on the global community.

Moreover, afforestation appears to not provide the same benefits in all regions of the world. In particular, boreal forests appear to provide the smallest net benefits, as the carbon sequestrations benefits of conifer forests are mitigated by the reduction in albedo in winter [@mykleby2017] [@li2015]. Coniferous forests have also been shown to increase soil organic carbon less than broadleaf forests [@guo2021]. With tropical forests, a strong positive result is derived from afforestation, as the carbon sequestration benefit is complemented by a cooling impact due to changes in both albedo and evapotranspiration, and with temperate forests, the impact of afforestation is also shown to be positive, albeit less strongly so due to a winter warming effect [@li2015].

South America and Sub-Saharan Africa are two regions that show much potential for afforestation. It is estimated that these two regions hold at least 50% of the potential global gain while regions such as Northern Africa and the Middle East show little promise as forest growth rates are quite low [@doelman2020]. Thus, it is promising that the African Forest Landscape Restoration Initiative (AFR100), which aims to afforest 100 million hectares by 2030, has roughly 70% of its committments coming from Sub-Saharan nations [@auda2023]. While many of these nations may be well-suited in biophysical terms for afforestation, there is concern about how long-lasting any investment may be due to proximate zones of political instability. There are other nations that have a big potential for gains from afforestation, but they also face challenges. India, for example, has a great potential to increase its carbon sequestration through afforestation, but due to its masssive population and food security worries, its NDC does not entertain the idea of agricultural land being repurposed and instead primarily considers targeting wastelands that may not be well suited for afforestation [@amjath2019].

This project will assess the impact of various likely scenarios by melding together two widely accepted global models, the Global Trade Analysis Project (GTAP) computable general equilibrium model [@aguilar2022] and the Global Timber Model (GTM) [@sohngen2001]. The GTAP model has traditionally been used to analyze the impact of various shocks on things like agricultural prices and trade volumes. However, the standard model does not well account for land-use changes, and the GTAP Land Use and Land Cover (LULC) database has limitations due to its infrequent updates. The GTM is quite helpful in getting around these issues, as it is a dynamic optimization model that can examine the impact of afforestation or avoided deforestation on the carbon sink. By joining the two models and passing outputs back and forth between the models, any limitations of either model can be minimized and more robust results, therefore, can be attained.

@@

1) Full attainment of AFR100 pledges as part of Bonn Challenge: This says that there will be a 127.81 million hectare increase in forest area just in Sub-Saharan Africa by 2030 [@bonn2023]. To see the list of countries and the hectares pledged, please view Scenario 1 of the Excel document "Bonn Challenge-Africa".

2) Partial attainment of AFR100 pledges as part of Bonn Challenge: This scenario analyzes full attainment for nations with conflict severity level less than or equal to limited and 0% attainment for regions with a higher conflict severity rating. Using this measure, there would only be a 54.95 million hectare increase in forest area by 2030 [@bonn2023]  [@acled2023]. To see the list of countries and the hectares pledged, please view Scenario 2 of the Excel document "Bonn Challenge-Africa".

3) Full attainment of Latin America pledges as part of Bonn Challenge: This says that there will be a 33.64 million hectare increase in forest area by 2030 [@bonn2023]. To see the list of countries and the hectares pledged, please view the Latin America tab of the Excel document "Bonn Challenge-Other Regions".

4) Full attainment of tropical nations' pledges as part of Bonn Challenge: This says that there will be a 182.30 million hectare increase in forest area by 2030 [@bonn2023]. To see the list of countries and the hectares pledged, please view the Tropical Forest tab of the Excel document "Bonn Challenge-Other Regions". Nations that lie in both tropical and sub-tropical zones and taken to be tropical.

5) Full attainment of UN strategic plan for forests 2030 (UNFF): This says that all deforestation will halt and there will be a 3% (~120 million hectare) increase in forest area worldwide by 2030 [@sewell2020]. This will be modeled as a uniform 3% increase in forested area across all nations.

6) Expansion of UN strategic plan for forests: This says that all deforestation will halt and there will be a 15% (~600 million hectare) increase in forest area worldwide by 2030 [@sewell2020]. This will be modeled as a uniform 15% increase in forested area across all nations.

# References
